
# Data Preparation {#dataprep}

```{r setup, include=F}
#knitr::opts_chunk$set(eval = F)
knitr::opts_chunk$set(eval = T)
```

Prior to undertaking the analysis, we need to prepare our data.  We are looking at annual county-level statistics, so here we will create an observation for each countyXyear for all US counties (n = 3143) and all years between 2010 and 2017.  To replicate this code, the data has been made available in the [following repository]().

## CDC Wonder Data

### Upload the Datas

We will start this process by using data downloaded from the [CDC Wonder](https://wonder.cdc.gov/mcd.html) database.  Wonder allows users to extract annual, county-level overdose numbers overall and for specific drugs (ie heroin, cocaine, synthetic opioids, et al).  Here we upload datasets extracted from the CDC Wonder database to create our initial dataset.

```{r wonder data, warning=FALSE}

## load the data

synthetic_opioid_data <- read.table(file = "Data/Synthetic_Opioid_Overdose_By_County_2010-2017.txt", sep = '\t', header = TRUE)
heroin_data <- read.table(file = "Data/Heroin_Overdose_By_County_2010-2017.txt", sep = '\t', header = TRUE)
cocaine_data <- read.table(file = "Data/Cocaine_Overdose_By_County_2010-2017.txt", sep = '\t', header = TRUE)
meth_data <- read.table(file = "Data/Meth_Overdose_By_County_2010-2017.txt", sep = '\t', header = TRUE)

## clean the data a little bit

synthetic_opioid_data$County <- as.character(synthetic_opioid_data$County)
synthetic_opioid_data$Deaths <- (as.character(synthetic_opioid_data$Deaths))
synthetic_opioid_data$Deaths[synthetic_opioid_data$Deaths == "Suppressed"] <- -99
synthetic_opioid_data$Deaths <- (as.numeric(synthetic_opioid_data$Deaths))
synthetic_opioid_data$Population <- as.numeric(as.character(synthetic_opioid_data$Population))

heroin_data$County <- as.character(heroin_data$County)
heroin_data$Deaths <- (as.character(heroin_data$Deaths))
heroin_data$Deaths[heroin_data$Deaths == "Suppressed"] <- -99
heroin_data$Deaths <- (as.numeric(heroin_data$Deaths))
heroin_data$Population <- as.numeric(as.character(heroin_data$Population))

cocaine_data$County <- as.character(cocaine_data$County)
cocaine_data$Deaths <- (as.character(cocaine_data$Deaths))
cocaine_data$Deaths[cocaine_data$Deaths == "Suppressed"] <- -99
cocaine_data$Deaths <- (as.numeric(cocaine_data$Deaths))
cocaine_data$Population <- as.numeric(as.character(cocaine_data$Population))

meth_data$County <- as.character(meth_data$County)
meth_data$Deaths <- (as.character(meth_data$Deaths))
meth_data$Deaths[meth_data$Deaths == "Suppressed"] <- -99
meth_data$Deaths <- (as.numeric(meth_data$Deaths))
meth_data$Population <- as.numeric(as.character(meth_data$Population))

## merge into one file

data <- synthetic_opioid_data[c(2:4,7:6)]

data$heroin_deaths <- NA
data$cocaine_deaths <- NA
data$meth_deaths <- NA

for(county_id in unique(data$County.Code)){
  for(year in unique(data$Year)){
    
    data$heroin_deaths[data$County.Code == county_id & data$Year == year] <- heroin_data$Deaths[heroin_data$County.Code == county_id & heroin_data$Year == year]
    data$cocaine_deaths[data$County.Code == county_id & data$Year == year] <- cocaine_data$Deaths[cocaine_data$County.Code == county_id & cocaine_data$Year == year]
    data$meth_deaths[data$County.Code == county_id & data$Year == year] <- meth_data$Deaths[meth_data$County.Code == county_id & meth_data$Year == year]
  }
}

colnames(data) <- c("county_name","county_code","year","population","synthetic_opioid_deaths","heroin_deaths","cocaine_deaths","meth_deaths")

data <- data[!is.na(data$population),]

```

### Imputing Missing Data

The CDC suppresses any value beneath 10 (ie if a county had 6 overdose deaths in a given year, the CDC just reports suppressed) and so we need to find a way to impute this value.  The [CDC recommends](https://ajph.aphapublications.org/doi/abs/10.2105/AJPH.2014.301900) a procedure which we update.  Essentially for a given state $S$ in a given year $Y$, we take the number of overdoses $O$ in that year and subtract from it all of the non-suppressed county counts $o_i$.  The remaining number of deaths, $O - \sum{o_i}$ is then distributed to the "Suppressed" counties on a population basis. 

```{r imputation, warning=FALSE}

## load the state data

synthetic_opioid_data <- read.table(file = "Data/Synthetic_Opioid_Overdose_By_State_2010-2017.txt", sep = '\t', header = TRUE)
heroin_data <- read.table(file = "Data/Heroin_Overdose_By_State_2010-2017.txt", sep = '\t', header = TRUE)
cocaine_data <- read.table(file = "Data/Cocaine_Overdose_By_State_2010-2017.txt", sep = '\t', header = TRUE)
meth_data <- read.table(file = "Data/Meth_Overdose_By_State_2010-2017.txt", sep = '\t', header = TRUE)

## clean the data a little bit

synthetic_opioid_data$State <- as.character(synthetic_opioid_data$State)
synthetic_opioid_data$Deaths <- (as.character(synthetic_opioid_data$Deaths))
synthetic_opioid_data$Deaths[synthetic_opioid_data$Deaths == "Suppressed"] <- -99
synthetic_opioid_data$Deaths <- (as.numeric(synthetic_opioid_data$Deaths))

heroin_data$State <- as.character(heroin_data$State)
heroin_data$Deaths <- (as.character(heroin_data$Deaths))
heroin_data$Deaths[heroin_data$Deaths == "Suppressed"] <- -99
heroin_data$Deaths <- (as.numeric(heroin_data$Deaths))

cocaine_data$State <- as.character(cocaine_data$State)
cocaine_data$Deaths <- (as.character(cocaine_data$Deaths))
cocaine_data$Deaths[cocaine_data$Deaths == "Suppressed"] <- -99
cocaine_data$Deaths <- (as.numeric(cocaine_data$Deaths))

meth_data$State <- as.character(meth_data$State)
meth_data$Deaths <- (as.character(meth_data$Deaths))
meth_data$Deaths[meth_data$Deaths == "Suppressed"] <- -99
meth_data$Deaths <- (as.numeric(meth_data$Deaths))

## now we need to undertake the imputation procedure

for(state_id in unique(synthetic_opioid_data$State.Code)){
  for(year in unique(synthetic_opioid_data$Year)){
    
    ## we first want to get the total number of deaths for the state
    total_synthetic_opioid_deaths <- synthetic_opioid_data$Deaths[synthetic_opioid_data$State.Code == state_id & synthetic_opioid_data$Year == year]
    total_heroin_deaths <- heroin_data$Deaths[heroin_data$State.Code == state_id & heroin_data$Year == year]
    total_cocaine_deaths <- cocaine_data$Deaths[cocaine_data$State.Code == state_id & cocaine_data$Year == year]
    total_meth_deaths <- meth_data$Deaths[meth_data$State.Code == state_id & meth_data$Year == year]
    
    ## then we want to subtract the number of deaths recorded 
    ## and extract the counties that were suppressed
    
    county_data <- data[data$county_code >= state_id*1000 & data$county_code < (state_id*1000 + 1000) & data$year == year,]
    
    synthetic_opioid_counties_suppressed <- c()
    heroin_counties_suppressed <- c()
    cocaine_counties_suppressed <- c()
    meth_counties_suppressed <- c()
    
    for(county_id in unique(county_data$county_code)){
      
      ## synthetic opioid
      so_death <- county_data$synthetic_opioid_deaths[county_data$county_code == county_id]
      if(so_death == -99){
        synthetic_opioid_counties_suppressed <- c(synthetic_opioid_counties_suppressed, county_id)
      }else{
        total_synthetic_opioid_deaths <- total_synthetic_opioid_deaths - so_death
      }
      
      ## heroin
      h_death <- county_data$heroin_deaths[county_data$county_code == county_id]
      if(h_death == -99){
        heroin_counties_suppressed <- c(heroin_counties_suppressed, county_id)
      }else{
        total_heroin_deaths <- total_heroin_deaths - h_death
      }
      
      ## cocaine
      c_death <- county_data$cocaine_deaths[county_data$county_code == county_id]
      if(c_death == -99){
        cocaine_counties_suppressed <- c(cocaine_counties_suppressed, county_id)
      }else{
        total_cocaine_deaths <- total_cocaine_deaths - c_death
      }
      
      ## meth
      m_death <- county_data$meth_deaths[county_data$county_code == county_id]
      if(m_death == -99){
        meth_counties_suppressed <- c(meth_counties_suppressed, county_id)
      }else{
        total_meth_deaths <- total_meth_deaths - m_death
      }
    }
    
    
    ## now we get the population rate of deaths for suppressed counties
    synthetic_opioid_suppressed_pop <- 0
    heroin_suppressed_pop <- 0
    cocaine_suppressed_pop <- 0
    meth_suppressed_pop <- 0
    
    for(county_id in synthetic_opioid_counties_suppressed){
      synthetic_opioid_suppressed_pop <- synthetic_opioid_suppressed_pop + county_data$population[county_data$county_code == county_id]
    }
    for(county_id in heroin_counties_suppressed){
      heroin_suppressed_pop <- heroin_suppressed_pop + county_data$population[county_data$county_code == county_id]
    }
    for(county_id in cocaine_counties_suppressed){
      cocaine_suppressed_pop <- cocaine_suppressed_pop + county_data$population[county_data$county_code == county_id]
    }
    for(county_id in meth_counties_suppressed){
      meth_suppressed_pop <- meth_suppressed_pop + county_data$population[county_data$county_code == county_id]
    }
    
    synthetic_opioid_rate <- 0
    if(synthetic_opioid_suppressed_pop > 0){
      synthetic_opioid_rate <- total_synthetic_opioid_deaths/synthetic_opioid_suppressed_pop
    }
    
    heroin_rate <- 0
    if(heroin_suppressed_pop > 0){
      heroin_rate <- total_heroin_deaths/heroin_suppressed_pop
    }
    
    cocaine_rate <- 0
    if(cocaine_suppressed_pop > 0){
      cocaine_rate <- total_cocaine_deaths/cocaine_suppressed_pop
    }
    
    meth_rate <- 0
    if(meth_suppressed_pop > 0){
      meth_rate <- total_meth_deaths/meth_suppressed_pop
    }
    
    if(synthetic_opioid_rate < 0){ synthetic_opioid_rate <- 0}
    if(heroin_rate < 0){ heroin_rate <- 0}
    if(cocaine_rate < 0){ cocaine_rate <- 0}
    if(meth_rate < 0){ meth_rate <- 0}
    
    ## we now want to apply these rates to calculate an imputed value for each suppressed county in the actual data set
    ## note that we know that the crude number does not exceed 9 and thus we use the min() function to cap the value at 9
    
    for(county_id in unique(county_data$county_code)){
     
      county_pop <- data$population[data$county_code == county_id & data$year == year]
      
      if(data$synthetic_opioid_deaths[data$county_code == county_id & data$year == year] == -99){
        data$synthetic_opioid_deaths[data$county_code == county_id & data$year == year] <- min(county_pop*synthetic_opioid_rate,9)
      }
      
      if(data$heroin_deaths[data$county_code == county_id & data$year == year] == -99){
        data$heroin_deaths[data$county_code == county_id & data$year == year] <- min(county_pop*heroin_rate,9)
      }
      
      if(data$cocaine_deaths[data$county_code == county_id & data$year == year] == -99){
        data$cocaine_deaths[data$county_code == county_id & data$year == year] <- min(county_pop*cocaine_rate,9)
      }
      
      if(data$meth_deaths[data$county_code == county_id & data$year == year] == -99){
        data$meth_deaths[data$county_code == county_id & data$year == year] <- min(county_pop*meth_rate,9)
      }
    }
  }
}

```


### Calculating Crude Rates

Now, with the imputed values we are then able to calculate the crude deaths rates of each county (per 100,000 people).  This is done by dividing the number of deaths by the size of population and then multiplying this value by 100,000.

```{r crude rates}

data$synthetic_opioid_crude_death_rate <- 100000*(data$synthetic_opioid_deaths/data$population)
data$heroin_crude_death_rate <- 100000*(data$heroin_deaths/data$population)
data$cocaine_crude_death_rate <- 100000*(data$cocaine_deaths/data$population)
data$meth_crude_death_rate <- 100000*(data$meth_deaths/data$population)

```

## American Community Survey

We would also like to add some demographic data, which is available from the ACS.  This dataset has been created by merging multiple ACS data files, the code for which is contained in the appendix.

```{r ACS}

ACS <- read.csv("Data/Updated_ACS_data.csv", header = T)

for(i in 1:nrow(ACS)){
  for(j in 1:ncol(ACS)){
    if(is.na(ACS[i,j])){
      ACS[i,j] <- -999
    }
  }
}


data$unemployment_rate <- NA
data$median_household_income <- NA
data$mean_household_income <- NA
data$proportion_poverty <- NA
data$proportion_uninsured <- NA
data$proportion_homes_no_vehicle <- NA
data$proportion_homeowners_35perc_income <- NA
data$proportion_renters_35perc_income <- NA
data$proportion_white <- NA
data$proportion_black <- NA
data$proportion_american_indian_alaska_native <- NA
data$proportion_asian <- NA
data$proportion_native_hawaiian_pacific_islander <- NA
data$proportion_male <- NA
data$proportion_high_school_or_greater <- NA
data$proportion_bachelors_or_greater <- NA

for(i in 1:nrow(data)){
  
if(data$county_code[i] %in% c(2201,2232,2280)){} ## three Alaskan counties which don't have data.
  else{
  
data$unemployment_rate[i] <- ACS$Unemployment_Rate[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$median_household_income[i] <- ACS$Median_Household_Income[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$mean_household_income[i] <- ACS$Mean_Household_Income[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_poverty[i] <- ACS$Proportion_Poverty[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_uninsured[i] <- ACS$Proportion_Uninsured[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_homes_no_vehicle[i] <- ACS$Proportion_Homes_No_Vehicle[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_homeowners_35perc_income[i] <- ACS$Proportion_Homeowners_35Perc_Income_on_Home[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_renters_35perc_income[i] <- ACS$Proportion_Renters_35Perc_Income_on_Rent[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_white[i] <- ACS$Proportion_White[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_black[i] <- ACS$Proportion_Black[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_american_indian_alaska_native[i] <- ACS$Proportion_American_Indian_Alaska_Native[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_asian[i] <- ACS$Proportion_Asian[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_native_hawaiian_pacific_islander[i] <- ACS$Proportion_Native_Hawaiian_Pacific_Islander[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_male[i] <- ACS$Proportion_Male[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_high_school_or_greater[i] <- ACS$Proportion_High_School_or_Greater[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
data$proportion_bachelors_or_greater[i] <- ACS$Proportion_Bachelors_Degree_or_Greater[ACS$ID == data$county_code[i] & ACS$Year == data$year[i]]
  }
  
}

for(i in 1:nrow(ACS)){
  for(j in 1:ncol(ACS)){
    if(ACS[i,j] == -999){
      ACS[i,j] <- NA
    }
  }
}

```


## Additional Variables

### Hepatitis C Mortality (CDC Wonder)

We used a very similar approach for this variable as for our specific drug use mortalities, using state level estimates to impute suppressed county numbers.

```{r hepatitis}
hepatitis_state_data <- read.table(file = "Data/Hepatitis_C_Mortality_By_State.txt", sep = '\t', header = TRUE)
hepatitis_county_data <- read.table(file = "Data/Hepatitis_C_Mortality_By_County.txt", sep = '\t', header = TRUE)

## clean the data a little bit

hepatitis_state_data$State <- as.character(hepatitis_state_data$State)
hepatitis_state_data$Deaths <- (as.character(hepatitis_state_data$Deaths))
hepatitis_state_data$Deaths[hepatitis_state_data$Deaths == "Suppressed"] <- -99
hepatitis_state_data$Deaths <- (as.numeric(hepatitis_state_data$Deaths))

hepatitis_county_data$County <- as.character(hepatitis_county_data$County)
hepatitis_county_data$Deaths <- (as.character(hepatitis_county_data$Deaths))
hepatitis_county_data$Deaths[hepatitis_county_data$Deaths == "Suppressed"] <- -99
hepatitis_county_data$Deaths <- (as.numeric(hepatitis_county_data$Deaths))
hepatitis_county_data$Population <- as.numeric(as.character(hepatitis_county_data$Population))

## now we need to undertake the imputation procedure

for(state_id in unique(hepatitis_state_data$State.Code)){
  for(year in unique(hepatitis_state_data$Year)){
    
    ## we first want to get the total number of deaths for the state
    total_hepatitis_deaths <- hepatitis_state_data$Deaths[hepatitis_state_data$State.Code == state_id & hepatitis_state_data$Year == year]
    
    ## then we want to subtract the number of deaths recorded 
    ## and extract the counties that were suppressed
    
    county_data <- hepatitis_county_data[hepatitis_county_data$County.Code >= state_id*1000 & hepatitis_county_data$County.Code < (state_id*1000 + 1000) & hepatitis_county_data$Year == year,]
    
    counties_suppressed <- c()

    
    for(county_id in unique(county_data$County.Code)){
      
      death <- county_data$Deaths[county_data$County.Code == county_id]
      if(is.na(death)){
        
      }
      else if(death == -99){
        counties_suppressed <- c(counties_suppressed, county_id)
      }else{
        total_hepatitis_deaths <- total_hepatitis_deaths - death
      }
    }
    
    
    ## now we get the population rate of deaths for suppressed counties
    suppressed_pop <- 0
    
    for(county_id in counties_suppressed){
      suppressed_pop <- suppressed_pop + county_data$Population[county_data$County.Code == county_id]
    }
    
    death_rate <- 0
    if(suppressed_pop > 0){
      death_rate <- total_hepatitis_deaths/suppressed_pop
    }
    
    if(death_rate < 0){ death_rate <- 0}
    
    
    ## we now want to apply these rates to calculate an imputed value for each suppressed county in the actual data set
    ## note that we know that the crude number does not exceed 9 and thus we use the min() function to cap the value at 9
    
    for(county_id in unique(county_data$County.Code)){
     
      county_pop <- county_data$Population[county_data$County.Code == county_id]
      
      if(is.na(county_data$Deaths[county_data$County.Code == county_id])){
        
      }else if(county_data$Deaths[county_data$County.Code == county_id] == -99){
        hepatitis_county_data$Deaths[hepatitis_county_data$County.Code == county_id & hepatitis_county_data$Year == year] <- min(county_pop*death_rate,9)
      }
    }
  }
}

## so now we want to merge this data into our final dataset
## note that we need to convert this into an actual rate when we do this

data$hep_c_mortality_rate <- NA

for(i in 1:nrow(data)){
  
  year <- data$year[i]
  county <- data$county_code[i]
  
  numbers_of_deaths <- hepatitis_county_data$Deaths[hepatitis_county_data$County.Code == county & hepatitis_county_data$Year == year]
  population <- data$population[i]
  
  hepatitis_mortality_rate <- 100000*(numbers_of_deaths/population)
  
  data$hep_c_mortality_rate[i] <- hepatitis_mortality_rate
  
}

```

### Urbanicity

```{r urbanicity}

urbanicity <- read.csv("Data/Urbanicity.csv", header = T)

data$urbanicity <- NA
for(id in unique(data$county_code)){
  
  urban <- urbanicity$X2013.code[urbanicity$ï..FIPS.code == id]
  data$urbanicity[data$county_code == id] <- urban
  
}

```

### Opioid Prescribing 

https://www.cdc.gov/drugoverdose/maps/rxrate-maps.html


```{r prescribing}

prescribing <- read.csv("Data/Opioid_Prescribing_per_100.csv", header = T)
prescribing$Rate_per_100 <- as.numeric(as.character(prescribing$Rate_per_100))

data$opioid_prescriptions_per_100 <- NA

for(i in 1:nrow(data)){
  year <- data$year[i]
  county <- data$county_code[i]
  
  prescriptions <- prescribing$Rate_per_100[prescribing$State.County.FIPS.Code == county & prescribing$Year == year]
  
  if(length(prescriptions) > 0){
    data$opioid_prescriptions_per_100[i] <- prescriptions
  }
}

### Need to Impute the missing data
### Lets match on state and urbanicity?

for(year in 2010:2017){
  for(state in 1:56){
    
    counties <- data[data$year == year & data$county_code >= state*1000 & data$county_code < (state + 1)*1000,]
    for(urbanicity in 1:6){
      
      urb_cnty <- counties[counties$urbanicity == urbanicity,]
      rate <- mean(urb_cnty$opioid_prescriptions_per_100, na.rm = T)
      
      data$opioid_prescriptions_per_100[data$urbanicity == urbanicity & data$county_code %in% urb_cnty$county_code & data$year == year & is.na(data$opioid_prescriptions_per_100)] <- rate
      
    }
    
  }
}


```



### Population Density


```{r pop dens}

land_area <- read.csv("Data/Land_Area.csv", header = T)

data$population_density <- NA
for(id in unique(data$county_code)){
  
  if(id %in% c(2201,2232,2280)){} ## three Alaskan counties which don't have data.
  else{
    
    la <- land_area$Land_Area[land_area$FIPS == id]
    
    
    
    subset <- data[data$county_code == id,]
    for(year in unique(subset$year)){
      
      population <- subset$population[subset$year == year]
      pop_density <- population/la
      data$population_density[data$county_code == id & data$year == year] <- pop_density
      
    }
  }
  
  
}

data$log_population_density <- log(data$population_density)

```


### Police Violence

```{r police violence}

police_violence <- read.csv("Data/annual_police_violence_by_county.csv", header = T)

data$police_violence <- NA

for(i in 1:nrow(data)){
  
  year <- data$year[i]
  fips <- data$county_code[i]
  
  if(year == 2010){
    
    killings <- police_violence$deaths_2010[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
    
  }else if(year == 2011){
    killings <- police_violence$deaths_2011[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2012){
    killings <- police_violence$deaths_2012[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2013){
    killings <- police_violence$deaths_2013[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2014){
    killings <- police_violence$deaths_2014[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2015){
    killings <- police_violence$deaths_2015[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2016){
    killings <- police_violence$deaths_2016[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }else if(year == 2017){
    killings <- police_violence$deaths_2017[police_violence$county_fips == fips]
    
    if(length(killings) > 0 && killings > 0){
      data$police_violence[i] <- T    
    }else if(length(killings) >0 && killings == 0){
      data$police_violence[i] <- F 
    }
  }
}


```

### Road Access & Urgent Care Facilities

```{r roads and urgent care}

hospitals <- read.csv("Data/roads_urgent_care.csv", header = T)

data$road_access <- NA
data$urgent_care <- NA

for(id in unique(data$county_code)){
  
  road_access <- hospitals$roads_access[hospitals$GEOID == id]
  urgent_care <- hospitals$urgent_care[hospitals$GEOID == id]
  
  if(length(road_access) > 0 && road_access == "yes"){
    road_access <- T
  } else{ road_access<- F}
  
  if(length(urgent_care) > 0 && urgent_care == "yes"){
    urgent_care <- T
  }else{urgent_care <- F}
  
  data$road_access[data$county_code == id] <- road_access
  data$urgent_care[data$county_code == id] <- urgent_care
}


```

## Defining Outcomes

So, we have established the annual data, but, because we are looking at projecting into the future, we need to pair observations with future outcomes.  Essentially, we will pair each observation with the synthetic opioid death rates corresponding to one year, two years, and three years into the future.  This is of interest because mortality data isn't typically available in time to make next year predictions.  We can use these to show the importance of releasing data swiftly.


```{r outcomes}

## for now we are focusing on synthetic opioids
## we will create two variables, next year and in two years

data$next_year_synthetic_opioid_death_rate <- NA
data$two_year_out_synthetic_opioid_death_rate <- NA

for(fips in unique(data$county_code)){
  
  for(year in 2010:2016){
    
    next_year <- data$synthetic_opioid_crude_death_rate[data$year == year + 1 & data$county_code == fips]
    if(length(next_year) > 0){
      data$next_year_synthetic_opioid_death_rate[data$year == year & data$county_code == fips] <- next_year
    }
    if(year < 2016){
      second_year <- data$synthetic_opioid_crude_death_rate[data$year == year + 2 & data$county_code == fips]
      if(length(second_year) > 0){
      data$two_year_out_synthetic_opioid_death_rate[data$year == year & data$county_code == fips] <- second_year
      }
    }
  }
}

```

## Neighboring County Variables (including Gravity)

Of interest is considering how the presence of overdose in neighboring counties impacts overdose in a given county.  A gravity approach is considered appropriate so here we will create gravity metrics for each of the four classes of overdose we have defined in our dataset (synthetic opioids, heroin, meth, cocaine).  We multiply the overdose rates of two counties and divide by the distance to get the gravity between two counties.  Each county's gravity metric is the sum of all gravity scores.


```{r gravity, eval=F}
# rm(list=setdiff(ls(), "data"))
# distances <- read.csv("Data/County_Distances.csv", header = T)
# 
# data$synthetic_opiod_gravity <- NA
# data$heroin_gravity <- NA
# data$cocaine_gravity <- NA
# data$meth_gravity <- NA
# 
# for(i in 1:nrow(data)){
#   
#   print(i)
#   
#   fips <- data$county_code[i]
#   year <- data$year[i]
#   
#   synth_od <- data$synthetic_opioid_crude_death_rate[i]
#   heroin_od <- data$heroin_crude_death_rate[i]
#   cocaine_od <- data$cocaine_crude_death_rate[i]
#   meth_od <- data$meth_crude_death_rate[i]
#   
#   dist <- distances[distances$county1 == fips & distances$mi_to_county < 200,]
#   
#   synth_gravity <- 0
#   heroin_gravity <- 0
#   cocaine_gravity <- 0
#   meth_gravity <- 0
#   
#   for(j in 1:nrow(dist)){
#     
#     fips2 <- dist$county2[j]
#     d <- dist$mi_to_county[j]
#     
#     synth_od2 <- data$synthetic_opioid_crude_death_rate[data$county_code == fips2 & data$year == year]
#     heroin_od2 <- data$heroin_crude_death_rate[data$county_code == fips2 & data$year == year]
#     cocaine_od2 <- data$cocaine_crude_death_rate[data$county_code == fips2 & data$year == year]
#     meth_od2 <- data$meth_crude_death_rate[data$county_code == fips2 & data$year == year]
#     
#     if(length(synth_od2) > 0){
#       synth_gravity <- synth_gravity + ((synth_od*synth_od2)/(d^2))
#     }
#     
#     if(length(heroin_od2) > 0){
#       heroin_gravity <- heroin_gravity + ((heroin_od*heroin_od2)/(d^2))
#     }
#     
#     if(length(cocaine_od2) > 0){
#       cocaine_gravity <- cocaine_gravity + ((cocaine_od*cocaine_od2)/(d^2))
#     }
#     
#     if(length(meth_od2) > 0){
#       meth_gravity <- meth_gravity + ((meth_od*meth_od2)/(d^2))
#     }
#     
#   }
#   
#   
#   data$synthetic_opiod_gravity[i] <- synth_gravity
#   data$heroin_gravity[i] <- heroin_gravity
#   data$cocaine_gravity[i] <- cocaine_gravity
#   data$meth_gravity[i] <- meth_gravity
#   
# }

```

## NFLIS

```{r NFLIS}
rm(list=setdiff(ls(), "data"))
NFLIS <- read.csv("Data/NFLIS_by_Year.csv", header = T)
NFLIS <- NFLIS[1:51,]

data$state <- floor(data$county_code/1000)
data$NFLIS <- NA

for(i in 1:nrow(data)){
  
  year <- data$year[i]
  state <- data$state[i]
  
  total <- NA
  if(year == 2010){
    total <- NFLIS$NFLIS_2010[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2011){
    total <- NFLIS$NFLIS_2011[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2012){
    total <- NFLIS$NFLIS_2012[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2013){
    total <- NFLIS$NFLIS_2013[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2014){
    total <- NFLIS$NFLIS_2014[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2015){
    total <- NFLIS$NFLIS_2015[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2016){
    total <- NFLIS$NFLIS_2016[NFLIS$ï..STATE_FIP == state]
  }else if(year == 2017){
    total <- NFLIS$NFLIS_2017[NFLIS$ï..STATE_FIP == state]
  }
  
  data$NFLIS[i] <- total
}


```

## Save Dataset

Once completed, we then want to save this dataset for future use.  Creating this data (in particular, the gravity variables) is time consuming and so we do not want to have to frequently re-run this code.

```{r save}

rm(list=setdiff(ls(), "data"))
write.csv(data, "Final_Dataset.csv")

```